Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-06 MScDiplom_Lemenkova.tex
Comment 15 of Valentijn: MERIS sensor (Fig.4.8)
Comment 8 of Valentijn: discussed spectra of P.oceanica with Thalassia and other seagrasses (p.2.5.3)

\subsection{Seagrass global monitoring: history and perspectives}\index[concepts]{Seagrasses!global monitoring}
Mapping and monitoring the seagrass is important for the environmental assessment of the marine
ecosystems in coastal areas. Regular tracking of current distribution of seagrass meadows, based on
correct information and cartographic visualization of seagrasses, is a preventive environmental
management, which helps to analyse potential environmental risks of coastal areas, decrease of the
number of species, loss of meadows and patches of the seagrasses.
The tradition of global seagrass mapping though has not a very long history comparing to the
terrestrial cartography, due to the technical difficulties of underwater observations, yet nowadays is
has become a successful, rapidly developing and increasingly popular and challenging research branch.\\
Seagrasses can be mapped using various approaches, with the most popular method of the remote
sensing application, enabling to detect and discriminate the substratum and various vegetation types
growing on the seafloor.\\
Several global seagrass surveys\index[concepts]{Seagrasses!global surveys} organize and provide regular monitoring of the seagrass distribution,
health and species sustainability, amongst which the most world-known are, perhaps, the following
ones: \href{http://www.seagrassnet.org/}{Global Seagrass Monitoring project}, \href{http://www.seagrasswatch.org/}{Seagrasswatch}, 
located in Australia, two European-located institutions - \href{http://mediterranean.seagrassonline.org/}{the Mediterranean association Seagrass-2000} and \href{http://www.imedea.uib.es/index.php}{the Mediterranean Institute for Advanced Studies}, US American 
\href{http://www2.fiu.edu/~seagrass/}{Seagrass Ecosystems Research Laboratory in South Florida} and
\href{http://www.flseagrass.org/}{Florida Seagrass organisation}, the \href{http://www.seagrasses.org/}{Seagrasses.org} and the \href{http://wsa.seagrassonline.org/}{World Seagrass Association}.
These organizations aim at the global seagrass monitoring, providing with
research results and reporting guidelines and manuals with standardized methods and
recommendations, specific for the seagrass research and monitoring. There are also university marine
centres and research institutes conducting seagrass monitoring and as a particular part of their
research. Their reports and guidelines were used for references in the current research.\\
Regular observations and monitoring of the seagrasses are known since 1960s, mainly in tropical
regions (Australia). Since that time traditional methods of the seagrass monitoring and common
recommendations are being elaborated. The development of the underwater SCUBA diving
equipment and devices enabled to conduct underwater detailed measurements and observations
largely contributed to the improvement of the traditional \textit{in-situ} observations of seagrass. From the
other side, development of the remote sensing methods and data acquisition from space contributed to
the new methods of seagrass mapping, using distance approach and generally based on images
classification.

\subsection[Measuring water optical properties]{Measuring water optical properties: \\hyperspectral radiometers}
The application of the remote sensing data for seagrass mapping is based on our knowledge of the
spectral reflectance properties of the target objects, and using it for the classification of these objects
on the image. In case of seagrasses it is spectral reflectance of the seafloor cover types, which can be
analyzed using measurements of optical properties of sea water: radiance and irradiance.
Optical remote sensing methods can get through the clear waters to approximately 15–30 m \cite{Mumby04}\label{Mumby04}.
When sunlight enters the waters and goes down into the water column, parts of the
electromagnetic energy are absorbed and scattered, which is determined by the optical and physical
properties of the water, e.g. concentration of suspended particles, chlorofill, colored dissolved
organic matter (gelbstoff) that make up the water content \cite{Ackleson2003}\label{Ackleson, 2003}. Besides, light is strongly
dependent on wavelengths, i.e. it is greater in blue wavelengths (400 nm) than in others.

\begin{figure}
	\centering
	\subfloat [RAMSES-ACC-UV - Hyperspectral UVA/UVB Irradiance Sensor: 280-500 nm]{\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-2-1.jpg}}
	\subfloat [RAMSES-ARC - Hyperspectral UV-VIS Radiance Sensor: 320-950 nm.]{\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-2-2.jpg}}
	\caption{RAMSES Hyperspectral Sensors. Source: \href{http://www.trios.de}{Trios}}
	\label{fig:2.1}
\end{figure}

@@

		\item [*] An \ac{4AOP}\index[concepts]{Seawater optical properties!RTM!4AOP} is a French RTM freeware, designed in cooperation by \href{http://www.noveltis.fr/fr/index.php}{Noveltis} and \href{http://www.cnes.fr/}{the French Space Agency} (\ac{CNES}). Being a fine program which enables accurate computations of the radiance, defining other emissivity functions and other set up, it is however, particularly designed for the infrared waves (thus, not quite suitable in our case). To the advantages of this freeware we would count its user friendly interface and open source accessibility for the academic purposes. 

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-11.jpg}
	\caption{\textit{In situ} optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}}
	\label{fig:13}
\end{wrapfigure}

@@

Thus, the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, \ac{Landsat TM}, and \ac{CASI}\index{concepts}{Remote sensing!multispectral imagery!CASI}, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and \ac{CASI}, while \ac{Landsat TM} has not high
enough resolution. 
It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that \ac{CASI} enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates high suitability of \ac{CASI} images for the fine-scale mapping of the seagrass
landscapes. Thus, \ac{CASI} and Quickbird-2\index[concepts]{Remote sensing!multispectral imagery!Quickbird} images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.
\ac{ASTER}\index[concepts]{Remote sensing!multispectral imagery!ASTER}), launched in
1999 on board Terra sensor\index[concepts]{Remote sensing!multispectral imagery!Terra sensor}, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:2.4}). It has diverse
subsystems for the \ac{VNIR} with 15 m resolution, shortwave infrared \ac{SWIR}, and
\ac{TIR} wavelength regions \cite{Kalinowski04}\label{Kalinowski04}. For each channel there is separate
on board calibration \ac{OBC} system, telescope with independent pointing and different detector technology \cite{Arai05}\label{Arai05}. 
This makes \ac{ASTER} imagery especially suitable for the application towards detailed
mapping of surface temperature, emissivity and reflectance of objects and bathymetric elevations as
well. Successful application of the \ac{ASTER} imagery towards mapping of submerged vegetation
reported, for example, by Hirose \cite{Hirose04}\label{Hirose04}. 
Other multispectral images have also been used for the
seagrass interpretation. For example, the success of using imagery from the multispectral airborne
scanner Daedalus AADS1268 is reported by Heege et al \cite{Heege03}\label{Heege03} where they aim at classification of
macrophytes in shallow waters of the Lake Constance.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-15.jpg}
	\caption{Reflectance spectra of seagrass; Rrs(0-) - subsurface RS reflectance; Rrs b - the bottom
		reflectance. Source:\cite{Yang10}\label{Yang10}}
	\label{fig:2.6}
\end{wrapfigure}

In regard to the methods chosen for the image interpretation, the supervised classifications proves to
be the most worthy in a majority of case studies \cite{Palandro03,Peneva08}\label{Palandro03}\label{Peneva08}. While
often methods of the unsupervised classification are used as a tool for classifying submerged object
and features on the multispectral images and aerial imagery \cite{Fletcher09}\label{Fletcher09}, it still does not
provide common algorithms that can be applied to other images and regions. Thus, the strength of the
optical signal coming from the various types of the seafloor is strongly influenced by the effects of the
water column, its depth and chemical properties. Yet methods of the unsupervised classification
overleap mixed spectral effects of the water column which shift the real values of the spectra, with the
pure reflectance from the benthos; as a result, it may cause significant errors \cite{Dierssen03}\label{Dierssen03}.
Therefore, for the accurate assessments of various seafloor cover types, water
column depth and its optical properties, methods of supervised classification should be preferably
used as being more suitable in classification and interpretations of imagery.
While this and other studies \cite{Phinn08}\label{Phinn08} demonstrated the advantages and success of the
application of multispectral imagery for the spectral discrimination of the seafloor cover types and
mapping the submerged landscapes on the basis of pixels classification, the application of data from
hyperspectral sensors has better potential due to their higher resolution (Fig.\ref{fig:2.5}).

@@

@@

initial data of measurements of the spectral reflectance have been preprocessed and statistically analyzed thereafter with graphs visualizing mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:4.10}, \ref{fig:28}). These data
have been measured using different step of the wavebands: some measurements

@@

Therefore, after several experiments with
various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex
hulls made on its control points, and therefore is best suitable for our case: analysis of optical

@@

\begin{figure}[H]
	\centering
	\subfloat [depth of 0.5 m]{\label{fig:32a}\includegraphics[width=0.4\textwidth]{Fig-32a.jpg}}
	\subfloat [depth of 2.0 m]{\label{fig:32b}\includegraphics[width=0.4\textwidth]{Fig-32b.jpg}}\\
	\subfloat [depth of 3.5 m]{\label{fig:32c}\includegraphics[width=0.4\textwidth]{Fig-32c.jpg}}
	\subfloat [depths of 0.5, 2.0 and 3.5 m]{\label{fig:32d}\includegraphics[width=0.4\textwidth]{Fig-32d.jpg}}
	\caption{Remote sensing reflectance of \textit{P.oceanica} at various depths}
	\label{fig:4.9}
\end{figure}

@@

\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.35]{Sampling_design.jpg}
		\caption{Scheme of the depths measurements; fieldwork on Ligaria beach}\label{fig:A.4}
	\end{center}	
\end{figure}
\pagebreak

@@

	\vspace{1ex}
	\begin{tabular}{lrlr}
		Mean dependent var &  0.136154 & S.D. dependent var &  0.149193 \\
		Sum squared resid &  5.689669 & S.E. of regression &  0.247344 \\
	\end{tabular}
	\label{tab:27}
	\end{center}
\end{table}
%%%Gretl end
\pagebreak

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-2.jpg}
		\caption{The point querying shows the selected points and their coordinates within the area of seagrass meadow (green). \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.39}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-3.jpg}
	\caption{\textit{Logarithmic Enhancement to Raster}, applied to the aerial Google Earth image: the seagrass meadow can now be easily seen as a bright spot of purple color. \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.40}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{OpenEV-1.jpg}
		\caption{Raster properties dialog: visualization and spatial info about the image (projection UTM, zone 35, datum WGS-84, etc.)  \href{http://openev.sourceforge.net/}{OpenEV})}\label{fig:A.41}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Ilwis-2.jpg}
		\caption{Color composite image composed of 3 images of Cretan shelf, Google Earth. \href{http://www.ilwis.org/}{ILWIS}}\label{fig:A.42}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Fig-39.jpg}
		\caption{Results of the unsupervised classification of the seafloor cover types and land structure, Agia Pelagia; raster layer read into the \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS} project}
		\label{fig:42}
	\end{center}
\end{figure}
\begin{figure}
	\begin{center}
		\includegraphics[scale=0.40]{Fig-36.jpg}
		\caption{Results of the image classification in \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: \\ seagrass distribution in Bali area, Crete. }
		\label{fig:39}
	\end{center}
\end{figure}
\pagebreak

>>>>

\subsection{Seagrass global monitoring: history and perspectives}\index[concepts]{Seagrasses!global monitoring}
Mapping and monitoring the seagrass is important for the environmental assessment of the marine
ecosystems in coastal areas. Regular tracking of current distribution of seagrass meadows, based on
correct information and cartographic visualization of seagrasses, is a preventive environmental
management, which helps to analyse potential environmental risks of coastal areas, decrease of the
number of species, loss of meadows and patches of the seagrasses.
The tradition of global seagrass mapping though has not a very long history comparing to the
terrestrial cartography, due to the technical difficulties of underwater observations. However, nowadays is
has become a rapidly developing, increasingly popular and challenging research branch.\vspace{1ex}
Regular observations and monitoring of the seagrasses are known since 1960s, mainly in tropical
regions (Australia). Since that time traditional methods of the seagrass monitoring and common
recommendations are being elaborated. The development of the underwater SCUBA diving
equipment and devices enabled to conduct underwater detailed measurements and observations
largely contributed to the improvement of the traditional \textit{in-situ} observations of seagrass. From the
other side, development of the remote sensing methods and data acquisition from space contributed to
the new methods of seagrass mapping, using distance approach and generally based on images classification.\vspace{1ex}\\
Various global seagrass survey organizations \index[concepts]{Seagrasses!global surveys} organize and provide regular monitoring of the seagrass species distribution, health and environmental sustainability. \\ We list below the most known seagrass research institutions: 
\begin{itemize}
	\item Global-scaled: \href{http://www.seagrassnet.org/}{Global Seagrass Monitoring Network} and the \href{http://wsa.seagrassonline.org/}{World Seagrass Association}; \\ The \href{http://www.unep-wcmc.org/marine/seagrassatlas/contents.htm}{World Atlas of Seagrasses} is published by the \href{http://www.unep.org/}{UNEP}.
	\item Australian \href{http://www.seagrasswatch.org/}{Seagrasswatch} (perhaps, the best organization, regularly publishing informative reports)
	\item European: \href{http://mediterranean.seagrassonline.org/}{the Mediterranean association Seagrass-2000}, \href{http://www.imedea.uib.es/index.php}{the Mediterranean Institute for Advanced Studies} and \href{http://www.seagrasses.org/}{Seagrasses.org};
	\item  US American \href{http://www.columbia.com/Seagrass-Recovery/About_Us_Giving_Back_Seagrass,default,pg.html}{seagrass recovery campaign} by the \href{http://www.seagrassgrow.org/}{Seagrassgrow}, \href{http://www2.fiu.edu/~seagrass/}{Seagrass Ecosystems Research Laboratory in South Florida}, \href{http://counties.cce.cornell.edu/suffolk/habitat_restoration/seagrassli/index.html}{Seagrass.LI} and
\href{http://www.flseagrass.org/}{Florida Seagrass organisation}; 
	\item Asian: \href{http://www.unepscs.org/}{UNEP/GEF South China Sea Project}, \href{http://www.marineconservationcambodia.org/Marine-Species-Photographic-Database/Cambodian-Marine-Life/Cambodian-Marine-Plants/Seagrasses.html}{Marine Conservation Cambodia} and \href{http://www.sosmalaysia.org/home.html}{Sosmalaysia.org}.
\end{itemize}
All these organizations aim at the global seagrass monitoring, providing with
research results and reporting guidelines and manuals with standardized methods and
recommendations, specific for the seagrass research and monitoring. There are also university marine
centres and research institutes conducting seagrass monitoring and as a particular part of their
research and reporting various approaches for the monitoring and mapping of the seagrasses, including remote
sensing applications. Their reports and guidelines were used for references in the current research.

\subsection[Measuring water optical properties]{Measuring water optical properties: \\hyperspectral radiometers}
Among various research approaches for the seagrass mapping the most popular methods are based on application of the remote sensing techniques,
 enabling to detect and discriminate the substratum and various vegetation types growing on the seafloor.
The application of the remote sensing data for seagrass mapping is based on our knowledge of the
spectral reflectance properties of the target objects, and using it for the classification of these objects
on the image. In case of seagrasses it is spectral reflectance of the seafloor cover types, which can be
analyzed using measurements of optical properties of sea water: radiance and irradiance.
Optical remote sensing methods can get through the clear waters to approximately 15–30 m \cite{Mumby04}\label{Mumby04}.
When sunlight enters the waters and goes down into the water column, parts of the
electromagnetic energy are absorbed and scattered, which is determined by the optical and physical
properties of the water, e.g. concentration of suspended particles, chlorofill, colored dissolved
organic matter (gelbstoff) that make up the water content \cite{Ackleson2003}\label{Ackleson, 2003}. Besides, light is strongly
dependent on wavelengths, i.e. it is greater in blue wavelengths (400 nm) than in others.

\begin{figure}[H]
	\centering
	\subfloat [RAMSES-ACC-UV - Hyperspectral UVA/UVB Irradiance Sensor: 280-500 nm]{\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-2-1.jpg}}
	\subfloat [RAMSES-ARC - Hyperspectral UV-VIS Radiance Sensor: 320-950 nm.]{\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-2-2.jpg}}
	\caption{RAMSES Hyperspectral Sensors. Source: \href{http://www.trios.de}{Trios}}
	\label{fig:2.1}
\end{figure}

@@

		\item [*] An \ac{4AOP}\index[concepts]{Seawater optical properties!RTM!4AOP} is a French RTM freeware, designed in cooperation by \href{http://www.noveltis.fr/fr/index.php}{Noveltis} and \href{http://www.cnes.fr/}{the French Space Agency}, \ac{CNES}. Being a fine program which enables accurate computations of the radiance, defining other emissivity functions and other set up, it is however, particularly designed for the infrared waves (thus, not quite suitable in our case). To the advantages of this freeware we would count its user friendly interface and open source accessibility for the academic purposes. 

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-11.jpg}
	\caption{\textit{In situ} optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}}
	\label{fig:13}
\end{wrapfigure}

@@

Thus, the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, \ac{Landsat TM}, and \ac{CASI}\index{concepts}{Remote sensing!multispectral imagery!CASI}, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and \ac{CASI}, while \ac{Landsat TM} has not high
enough resolution. 
It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that \ac{CASI} enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates high suitability of \ac{CASI} images for the fine-scale mapping of the seagrass
landscapes. Thus, \ac{CASI} and Quickbird-2\index[concepts]{Remote sensing!multispectral imagery!Quickbird} images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.
\ac{ASTER}\index[concepts]{Remote sensing!multispectral imagery!ASTER}), launched in
1999 on board Terra sensor\index[concepts]{Remote sensing!multispectral imagery!Terra sensor}, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:2.4}). It has diverse
subsystems for the \ac{VNIR} with 15 m resolution, shortwave infrared \ac{SWIR}, and
\ac{TIR} wavelength regions \cite{Kalinowski04}\label{Kalinowski04}. For each channel there is separate
on board calibration \ac{OBC} system, telescope with independent pointing and different detector technology \cite{Arai05}\label{Arai05}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-15.jpg}
	\caption{Reflectance spectra of \textit{Thalassia} seagrass; Rrs(0-) - subsurface RS reflectance; Rrs b - the bottom
		reflectance. Source:\cite{Yang10}\label{Yang10}}
	\label{fig:2.6}
\end{wrapfigure}

This makes \ac{ASTER} imagery especially suitable for the application towards detailed
mapping of surface temperature, emissivity and reflectance of objects and bathymetric elevations as
well. Successful application of the \ac{ASTER} imagery towards mapping of submerged vegetation
reported, for example, by Hirose \cite{Hirose04}\label{Hirose04}. 
Other multispectral images have also been used for the
seagrass interpretation. For example, the success of using imagery from the multispectral airborne
scanner Daedalus AADS1268 is reported by Heege et al \cite{Heege03}\label{Heege03} where they aim at classification of
macrophytes in shallow waters of the Lake Constance. 
In regard to the methods chosen for the image interpretation, the supervised classifications proves to
be the most worthy in a majority of case studies \cite{Palandro03,Peneva08}\label{Palandro03}\label{Peneva08}. While
often methods of the unsupervised classification are used as a tool for classifying submerged object
and features on the multispectral images and aerial imagery \cite{Fletcher09}\label{Fletcher09}, it still does not
provide common algorithms that can be applied to other images and regions. Thus, the strength of the
optical signal coming from the various types of the seafloor is strongly influenced by the effects of the
water column, its depth and chemical properties. Yet methods of the unsupervised classification
overleap mixed spectral effects of the water column which shift the real values of the spectra, with the
pure reflectance from the benthos; as a result, it may cause significant errors \cite{Dierssen03}\label{Dierssen03}.
Therefore, for the accurate assessments of various seafloor cover types, water
column depth and its optical properties, methods of supervised classification should be preferably
used as being more suitable in classification and interpretations of imagery.
While this and other studies \cite{Phinn08}\label{Phinn08} demonstrated the advantages and success of the
application of multispectral imagery for the spectral discrimination of the seafloor cover types and
mapping the submerged landscapes on the basis of pixels classification, the application of data from
hyperspectral sensors has better potential due to their higher resolution (Fig.\ref{fig:2.5}).

@@

\subsubsection{Comparison of seagrass spectra }\label{sec:2.5.3}

The reflectance spectra from different seagrass species reported in various research works demonstrate diverse spectral signatures which are dependent on the mixture of environmental conditions as well as individual characteristics of the seagrass. 
In the current work we have compiled an extensive measurements dataset (up to 400 single spectral profiles) of the reflectance spectra from various seafloor cover types (P.oceanica and carbonate sand) in the coastal waters of Crete Island, which is partly visualized on the plotted graphs (Fig.\ref{fig:4.2}, Fig.\ref{fig:4.10}, \ref{fig:A.12}, \ref{fig:A.13}, \ref{fig:A.15}, \ref{fig:A.25}, \ref{fig:A.26}).
Having analyzed reflectance spectra received in previous works by other researchers and compared them with our results, obtained in the current work, we noticed some general trends in the character of reflectance spectra of various seafloor cover types. We mention below some general conclusions and observed tendencies in the profiles and patterns of spectra of \textit{P.oceanica} and carbonate sand (as for Crete Island). 
Measured remote sensing spectral reflectance is highly variable with changing environmental conditions, such as water physical content (amounts of suspended particles and organic matter, salinity, chemical content), water column depth, atmospheric conditions (e.g. sun zenith angle) and individual features on the plant, e.g. morphology and colour pigmentation of the leaves, differed in various plant's health and age conditions. 
The maximal values of the spectral reflectance of \textit{P.oceanica}, received both in the current work and by previous researchers, lies in wavelength interval of 500-600 nm, as, for example, shown on Fig.\ref{fig:13}, Fig.\ref{fig:2.6}, and can be clearly seen in our results as well (Fig.\ref{fig:4.2}, Fig.\ref{fig:4.10}, Fig.\ref{fig:4.13}, \ref{fig:A.25}). Comparison of the remote sensing reflectance spectra of various seagrass species, e.g. \textit{P.oceanica} (Fig. \ref{fig:4.6}) and \textit{Thalassia testudinum} (Fig.\ref{fig:2.6}), shows that \textit{P.oceanica} is in generally brighter in the wavelength interval 500-600 nm which can be caused by various colour pigmentation. The plotted graphs with spectra of the seagrasses \textit{P.oceanica} and \textit{Thalassia testudinum} demonstrate that the dip in the spectral reflectance after ca 600 nm is more pronounced in case of \textit{P.oceanica} seagrass beds (Fig.\ref{fig:4.6}) than in those with \textit{Thalassia testudinum} (Fig.\ref{fig:2.6}), due to absorption caused by pigment of green leaves. 
The remote sensing of the bottom albedo of \textit{P.oceanica} and carbonate sand (Fig.\ref{fig:4.11}) shows clear difference between the carbonate sand and \textit{P.oceanica}, which is evidently much darker.  To estimate and further analyze particularities of the \textit{P.oceanica} remote sensing reflectance spectra over regions with different bathymetry (0.5, 1.5 and 2.5 meters column depth) we used \ac{WASI} \ac{RTM} software, which incorporates measurements of the seafloor reflectance and water column optical properties.

@@

initial data of measurements of the spectral reflectance have been preprocessed and statistically analyzed thereafter with graphs visualizing mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:4.10}, \ref{fig:28}). These data have been measured using different step of the wavebands: some measurements

@@

Therefore, after several experiments with various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex hulls made on its control points, and therefore is best suitable for our case: analysis of optical

@@

\begin{figure}[H]
	\centering
	\subfloat [depth of 0.5 m]{\label{fig:32a}\includegraphics[width=0.4\textwidth]{Fig-32a.jpg}}
	\subfloat [depth of 2.0 m]{\label{fig:32b}\includegraphics[width=0.4\textwidth]{Fig-32b.jpg}}\\
	\subfloat [depth of 3.5 m]{\label{fig:32c}\includegraphics[width=0.4\textwidth]{Fig-32c.jpg}}
	\subfloat [depths of 0.5, 2.0 and 3.5 m]{\label{fig:32d}\includegraphics[width=0.4\textwidth]{Fig-32d.jpg}}
	\caption{Remote sensing reflectance of \textit{P.oceanica} at various depths. MERIS sensor.}
	\label{fig:4.9}
\end{figure}

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.35]{Sampling_design.jpg}
		\caption{Scheme of the depths measurements; fieldwork on Ligaria beach}\label{fig:A.4}
	\end{center}	
\end{figure}
\pagebreak

@@

	\vspace{1ex}
	\begin{tabular}{lrlr}
		Mean dependent var &  0.136154 & S.D. dependent var &  0.149193 \\
		Sum squared resid &  5.689669 & S.E. of regression &  0.247344 \\
	\end{tabular}
	\label{tab:27}
	\end{center}
\end{table}
%%%Gretl end

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-2.jpg}
		\caption{The point querying shows the selected points and their coordinates within the area of seagrass meadow (green). \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.39}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-3.jpg}
	\caption{\textit{Logarithmic Enhancement to Raster}, applied to the aerial Google Earth image: the seagrass meadow seen as a bright spot of purple color. \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.40}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{OpenEV-1.jpg}
		\caption{Raster properties dialog: visualization and spatial info about the image (projection UTM, zone 35, datum WGS-84, etc.)  \href{http://openev.sourceforge.net/}{OpenEV})}\label{fig:A.41}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Ilwis-2.jpg}
		\caption{Color composite image composed of 3 images of Cretan shelf, Google Earth. \href{http://www.ilwis.org/}{ILWIS}}\label{fig:A.42}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Fig-39.jpg}
		\caption{Results of the unsupervised classification of the seafloor cover types and land structure, Agia Pelagia; raster layer read into the \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS} project}
		\label{fig:42}
	\end{center}
\end{figure}
\begin{figure}
	\begin{center}
		\includegraphics[scale=0.25]{Fig-36.jpg}
		\caption{Results of the image classification in \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: \\ seagrass distribution in Bali area, Crete. }
		\label{fig:39}
	\end{center}
\end{figure}
\pagebreak
